---
title: "Kalman Filter Pt. 1: Core Services vs NAIRU in Brazil"
description: "A from-scratch Kalman filter / MLE detour, then a three-package (MARSS, KFAS, kalmanfilter) replication of the BCB's time-varying NAIRU Phillips curve for services inflation."
date: 2026-07-09
categories: [kalman filter, brazil, nairu, inflation, time series]
---

## Motivation

Brazil's unemployment rate is at its lowest level in more than two decades
(around 6.5–7%), which usually means an overheating labor market. Yet
services inflation — arguably the part of the CPI most sensitive to labor
costs — is running at about 5% a year, well below the 8–9% seen back in
2014, the last time unemployment was this low. Why isn't tight labor supply
showing up in services prices the way it used to?

That question is the subject of a short 2025 post on the Central Bank of
Brazil's (BCB) research blog, ["O nível recente da inflação de serviços
deveria causar
surpresa?"](https://www.bcb.gov.br/noticiablogbc/29/noticia){target="_blank"}
by Thiago Trafane Oliveira Santos and Ricardo Sabbadini. Their answer hinges
on a single latent variable: the NAIRU (non-accelerating inflation rate of
unemployment) — the unemployment rate below which inflation tends to
accelerate — which they let vary over time via a Kalman filter. If the
NAIRU itself has fallen since 2018, a labor market that looks "hot" in raw
unemployment terms may not be as inflationary as it looks.

This post is **Pt. 1** of a two-part series. Here, I replicate the original
BCB exercise — same sample, same Phillips curve, same two-stage estimation
— in three different R packages (**MARSS**, **KFAS**, and **kalmanfilter**)
side by side, as a portfolio piece on applied Kalman filtering. I don't
update the data or add fresh economic analysis; that's **Pt. 2**.

Before touching the NAIRU model itself, it's worth building the Kalman
filter machinery generically — the state-space form, the predict/update
recursions, and how maximum likelihood estimation falls out of them almost
for free.

## The Kalman filter, generically

At its core, a Kalman filter is a running weighted average: at every time
step it has a guess about some quantity it can't observe directly (here,
the NAIRU), it sees a new, noisy clue about that quantity, and it blends
the old guess with the new clue — weighting each by how much it trusts it.
"State-space form" is just the bookkeeping that makes "how much to trust
each one" precise and automatic.

### State-space form

A linear-Gaussian state-space model has an **observation equation** linking
what we see, $y_t$, to a latent state $\alpha_t$:

$$
y_t = Z_t \alpha_t + \varepsilon_t, \qquad \varepsilon_t \sim N(0, H_t)
$$

and a **state (transition) equation** describing how the latent state
evolves:

$$
\alpha_t = T_t \alpha_{t-1} + R_t \eta_t, \qquad \eta_t \sim N(0, Q_t)
$$

with $\varepsilon_t$ and $\eta_t$ mutually independent at all leads and
lags. $Z_t$, $T_t$, $H_t$, $Q_t$ can depend on unknown parameters $\theta$
(and, if the model has covariates, on time) — that's exactly the machinery
this post's Phillips curve gets built on.

### Predict / update recursions

Given $\alpha_{t-1} \mid y_{1:t-1} \sim N(a_{t-1}, P_{t-1})$, the Kalman
filter produces $\alpha_t \mid y_{1:t} \sim N(a_t, P_t)$ in two steps.

**Predict** (propagate the state forward, before seeing $y_t$):

$$
a_{t|t-1} = T_t a_{t-1}, \qquad
P_{t|t-1} = T_t P_{t-1} T_t' + R_t Q_t R_t'
$$

**Update** (fold in the new observation):

$$
v_t = y_t - Z_t a_{t|t-1}, \qquad
F_t = Z_t P_{t|t-1} Z_t' + H_t
$$

$$
K_t = P_{t|t-1} Z_t' F_t^{-1}, \qquad
a_t = a_{t|t-1} + K_t v_t, \qquad
P_t = (I - K_t Z_t) P_{t|t-1}
$$

$v_t$ is the **prediction error** — the gap between what was observed and
what the model, given everything up to $t-1$, expected — and $F_t$ is its
variance. Everything the filter needs from the data at each step is
summarized by $(v_t, F_t)$.

### MLE via the prediction-error decomposition

The joint density of $y_1, \ldots, y_n$ factors as
$p(y_1)\prod_{t=2}^n p(y_t \mid y_{1:t-1})$, and because each conditional
density is exactly the Gaussian implied by $(v_t, F_t)$ above, the
log-likelihood collapses to a sum over the filter's own by-products — the
**prediction-error decomposition**:

$$
\log L(\theta) = -\frac{n}{2}\log(2\pi) - \frac{1}{2}\sum_{t=1}^n
\left( \log|F_t| + v_t' F_t^{-1} v_t \right)
$$

This is the entire reason the Kalman filter is an estimation tool and not
just a signal-extraction one: run the filter once for a candidate $\theta$,
read off $\log L(\theta)$ for free, and hand it to a numerical optimizer.
Every backend in this post — no matter how it likes to be *specified* —
ultimately reduces to this same loop.

## From generic theory to the BCB's NAIRU model

The BCB's Phillips curve for core services inflation is:

$$
\pi_t^{ss} = \alpha + \beta \pi_{t-1}^{12m} + (1-\beta) E_t(\pi^{LP})
  - \gamma (u_t - u_t^*) + \theta \, \text{IPPCV}_t + \varepsilon_t
$$

where $\pi_t^{ss}$ is core services inflation, $\pi_{t-1}^{12m}$ is the
prior month's 12-month headline IPCA (the inertial component),
$E_t(\pi^{LP})$ is the long-run Focus survey inflation expectation (the
expectational component), $u_t$ is the seasonally-adjusted unemployment
rate, $\text{IPPCV}_t$ is a supply-chain pressure index (zero outside
Oct/2020–Sep/2022), and $u_t^*$ — the NAIRU — is the one latent, time-varying
piece.

$\alpha$ (the average level gap between services-core and headline
inflation) is **calibrated**, not estimated — the mean of $\pi_t^{ss} -
\pi_t^{12m}$ over the pre-pandemic sample. Everything else is either a
known regressor or one of $\beta, \gamma, \theta, u_t^*$.

Mapping this onto the generic form: the observation equation is the
Phillips curve rewritten with $u_t^*$ isolated, and the state equation is a
random walk on the NAIRU itself:

$$
\pi_t^{ss} - \alpha - \beta \pi_{t-1}^{12m} - (1-\beta)E_t(\pi^{LP})
  - \theta\,\text{IPPCV}_t + \gamma u_t = \gamma u_t^* + \varepsilon_t
$$

$$
u_t^* = u_{t-1}^* + \vartheta_t, \qquad
\frac{\text{Var}(\varepsilon_t/\gamma)}{\text{Var}(\vartheta_t)} = \lambda
$$

This is the **"r-filter of order 1"** of Araujo, Rodrigues Neto & Areosa
(2003) — a Hodrick-Prescott-style generalization suited to series with no
growth trend (like an unemployment rate), calibrated at $\lambda=120$ (the
monthly analogue of the usual HP $\lambda=14{,}400$).

::: {.callout-note}
## A reparametrization worth flagging

$\gamma$ appears twice above — once as the loading on the latent state
$u_t^*$, once as the coefficient on the *known* regressor $u_t$ — which
none of this post's three packages can estimate directly (no built-in way
to tie a state loading to a regression coefficient elsewhere in the same
model). The fix is a change of state: let $s_t \equiv \sqrt{\lambda}\,
\gamma\, u_t^*$. The loading on $s_t$ becomes the **known constant**
$1/\sqrt{\lambda}$, its innovation variance turns out to equal
$\text{Var}(\varepsilon_t)$ **exactly** (no $\gamma$ left in the variance
ratio — the two powers of $\gamma$ cancel), and $\gamma$ is left doing only
one job: the coefficient on $u_t$. Every backend below is built on this
$s_t$, with $u_t^* = s_t / (\sqrt{\lambda}\,\gamma)$ recovered after
fitting.
:::

### Two-stage estimation

To keep the pandemic from distorting the structural parameters, the
original post estimates in two stages:

1. **Stage 1** (pre-pandemic window, IPPCV always zero): estimate $\beta$
   and $\gamma$.
2. **Stage 2** (full window, IPPCV included): **fix** $\beta$ and $\gamma$
   at their stage-1 values, estimate $\theta$ and the full NAIRU path.

Every backend below implements this through the same seam:
`fit_nairu(data, lambda, package, alpha, fixed)` — `fixed = NULL` for stage
1, `fixed = list(beta =, gamma =)` for stage 2.

This post only runs $\lambda=120$ (the original's central calibration) and
only the two-stage procedure. The [original
post](https://www.bcb.gov.br/noticiablogbc/29/noticia){target="_blank"}
also checks robustness at $\lambda=60$ and $\lambda=240$, and against a
single-stage estimation (no pandemic-window split) — both come out
qualitatively unchanged there, and aren't reproduced here.

## Data

::: {.callout-important}
## Deviation from the original: sample starts Jul/2006, not Dec/2001

The original post's sample runs Dec/2001–Nov/2024. This replication's core
services-core inflation series ($\pi_t^{ss}$) is only available from
**Jul/2006** onward in the data source used here — the underlying table
simply doesn't go back further. Everything else (unemployment, headline
IPCA, Focus expectations, IPPCV) does have data back to 2001 or earlier.
Per this project's qualitative-replication bar, the fix is to use the
longest available window and say so plainly rather than force a mismatch —
so stage 1 here runs Jul/2006–Dec/2019 (≈162 obs, vs. the original's 217),
and stage 2 runs Jul/2006–Nov/2024.
:::

The pipeline that assembles the raw series lives in `raw_data.R` and pulls
from the author's internal data infrastructure (`datalake.utils`) —
`read_vintage()` against a proprietary datalake, and `seas_dplyr()` for X13
seasonal adjustment. It's shown here **verbatim, but not executed** at
render time, both as an honest record of where the numbers come from and as
a visible nod to that infrastructure:

```{r}
#| label: raw-data-pipeline
#| eval: false
#| code-fold: true
#| code-summary: "raw_data.R (not executed — pulls from a proprietary datalake)"
#| file: "0001-kalman-filter-nairu-pt1/raw_data.R"
```

::: {.callout-note}
Because `read_vintage()` and `seas_dplyr()` need internal datalake access
and an X13 install, they won't run for anyone who clones this site's repo.
`raw_data.R` writes its output to `frozen_snapshot.csv`, which **is**
committed — that's the file every code chunk below actually reads, so
`quarto render` stays fully reproducible without depending on infra only I
have access to.
:::

```{r}
#| label: load-snapshot
#| code-fold: true
library(dplyr)
library(ggplot2)

source("0001-kalman-filter-nairu-pt1/R/fit_nairu.R")

db <- read.csv("0001-kalman-filter-nairu-pt1/data/frozen_snapshot.csv")
db$date <- as.Date(db$date)

stage1_data <- db %>% filter(date >= as.Date("2006-07-01"), date <= as.Date("2019-12-01"))
stage2_data <- db %>% filter(date >= as.Date("2006-07-01"), date <= as.Date("2024-11-01"))
alpha <- calibrate_alpha(stage1_data)
lambda <- 120

cat("alpha (calibrated):", round(alpha, 2), "p.p.\n")
cat("stage 1:", format(min(stage1_data$date)), "to", format(max(stage1_data$date)),
    "(", nrow(stage1_data), "obs )\n")
cat("stage 2:", format(min(stage2_data$date)), "to", format(max(stage2_data$date)),
    "(", nrow(stage2_data), "obs )\n")
```

$\alpha$ comes out close to the original's calibrated 1.1 p.p. — a useful
independent check that the underlying series (before any Kalman filtering)
are on the same footing as the original post's.

## Three packages, one model

Each package below fits the *same* two-stage model against the *same*
frozen snapshot, through the shared `fit_nairu()` contract. What differs is
how each package likes its state space handed to it — and, as a direct
consequence, how much of the parameter vector each one's own optimizer
searches over.

### MARSS: explicit matrices, one estimated element per cell

MARSS wants a fully-specified list of matrices (`B`, `U`, `Q`, `Z`, `A`,
`R`, plus `D`/`d` for regression effects), where any cell can be a fixed
number or a character label naming a free parameter — mix and match
freely within a matrix. It's the most literal translation of the algebra
above onto code.

The one thing MARSS's shorthand *doesn't* support is tying a free parameter
across two *different* top-level matrices — which is exactly what "$Q$
equals $R$" (this post's reparametrization) asks for. So `fit_nairu_marss()`
profiles the shared variance in an outer 1-D search, and lets MARSS's own
estimator handle $\beta,\gamma$ (or just $\theta$, in stage 2) conditional
on each candidate variance — the reason MARSS's runtime in the comparison
table below is visibly the largest of the three.

```{r}
#| label: marss-model
#| code-fold: false
#| eval: false
#| file: "0001-kalman-filter-nairu-pt1/R/fit_nairu_marss.R"
```

### KFAS: formulas, and regression coefficients concentrated out

KFAS builds models from a formula (`SSMtrend()`, `SSMregression()`), and
its exact diffuse initialization lets regression coefficients (here,
$\beta,\gamma$ or $\theta$) be **concentrated out of the likelihood
entirely** — treated as diffuse states with zero process variance, solved
via generalized least squares inside the same filter pass that estimates
the NAIRU. The only thing left for `fitSSM()`'s outer optimizer to search
over is the single shared variance parameter — which is why KFAS's runtime
below is the smallest of the three, despite estimating exactly the same
model.

The one manual step is overriding the trend component's default loading
(KFAS assumes 1) to this post's known $1/\sqrt{\lambda}$:

```{r}
#| label: kfas-model
#| code-fold: false
#| eval: false
#| file: "0001-kalman-filter-nairu-pt1/R/fit_nairu_kfas.R"
```

### kalmanfilter: bare matrices, roll your own optimizer

`kalmanfilter` is the most bare-bones of the three: one function,
`kalman_filter(ssm, yt)`, that runs the filter and hands back the
log-likelihood — no built-in estimator at all. Every parameter, structural
or variance, goes into a vector you optimize yourself (with `optim()`,
`maxLik`, or anything else); every state-space matrix is built from that
vector by hand, on every likelihood evaluation.

```{r}
#| label: kalmanfilter-model
#| code-fold: false
#| eval: false
#| file: "0001-kalman-filter-nairu-pt1/R/fit_nairu_kalmanfilter.R"
```

## Fitting all three

```{r}
#| label: fit-all
#| code-fold: true
packages <- c("marss", "kfas", "kalmanfilter")

stage1_fits <- setNames(
  lapply(packages, function(pkg) fit_nairu(stage1_data, lambda, pkg, alpha)),
  packages
)
stage2_fits <- setNames(
  lapply(packages, function(pkg) {
    fixed <- list(beta = stage1_fits[[pkg]]$beta, gamma = stage1_fits[[pkg]]$gamma)
    fit_nairu(stage2_data, lambda, pkg, alpha, fixed = fixed)
  }),
  packages
)
```

## Comparison

```{r}
#| label: tbl-comparison
#| code-fold: true
nairu_at <- function(fit, d) fit$nairu$nairu[fit$nairu$date == as.Date(d)]

comparison <- data.frame(
  Package = c("MARSS", "KFAS", "kalmanfilter", "Original (BCB, λ=120)"),
  beta = c(unname(sapply(stage1_fits, `[[`, "beta")), 0.30),
  gamma = c(unname(sapply(stage1_fits, `[[`, "gamma")), 0.67),
  theta = c(unname(sapply(stage2_fits, `[[`, "theta")), 1.32),
  `NAIRU@Nov2024` = c(unname(sapply(stage2_fits, nairu_at, "2024-11-01")), NA),
  logLik = c(unname(sapply(stage2_fits, `[[`, "loglik")), NA),
  `runtime (s)` = c(unname(sapply(stage2_fits, `[[`, "runtime_secs")), NA),
  check.names = FALSE
)
knitr::kable(comparison, digits = 3)
```

β and γ land close to the original's 0.30/0.67 in sign and order of
magnitude (γ running a bit higher here, consistent with the shorter,
more recent stage-1 window per the sample-start deviation above); θ lands
very close to the original's 1.32. The three packages agree with each
other closely — MARSS's nested profiling and KFAS/kalmanfilter's joint
optimization are different routes to (nearly) the same likelihood surface,
which is itself a useful cross-check that none of the three has a sign or
matrix-orientation bug. `logLik` isn't directly comparable across packages
here (each has its own convention for additive constants under diffuse
initialization) — read each package's value only relative to itself, e.g.
across a λ-robustness sweep, not across the table's rows.

```{r}
#| label: fig-nairu
#| fig-cap: "Estimated NAIRU path, Jul/2006–Nov/2024, all three packages"
#| code-fold: true
nairu_long <- bind_rows(lapply(names(stage2_fits), function(pkg) {
  df <- stage2_fits[[pkg]]$nairu
  df$package <- pkg
  df
}))

ggplot(nairu_long, aes(date, nairu, color = package)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 0, color = "grey80") +
  labs(x = NULL, y = "NAIRU (%)", color = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

::: {.zoomable-figure}
:::

The pattern matches the original post's own description: the NAIRU sits
persistently around 10% through 2018, then declines — landing close to
7.5% by Nov/2024 across all three packages, in the same ballpark as the
original's implied estimate (≈7.1–7.2%, back-calculated from its reported
unemployment gap of −1.6 to −1.7 p.p. against a Nov/2024 unemployment rate
of ≈5.4–5.5%).

## Takeaway

The Kalman filter's real trick isn't signal extraction — it's that the
same predict/update recursion that tracks a latent NAIRU also hands you its
log-likelihood for free, via the prediction-error decomposition. Once that
clicks, "which R package" becomes a question of how much of the parameter
vector you want to hand to that package's own optimizer versus solve
yourself — MARSS's explicit matrices, KFAS's diffuse-regression shortcut,
and kalmanfilter's roll-your-own filter are three different answers to
that question, applied here to the same Phillips-curve NAIRU model as the
BCB's own.

**Pt. 2** re-runs this same model on data extended to the latest available
vintage, and turns to the economic question the original post was actually
asking: with the NAIRU where it now sits, is Brazil's tight labor market
still surprising given today's services inflation? That means confidence
bands on the NAIRU path, an updated unemployment-gap read, and a look at
whether the Oct/2020–Sep/2022 IPPCV window still makes sense to extend.

## References

- Santos, T. T. O., & Sabbadini, R. (2025). [*O nível recente da inflação de
  serviços deveria causar
  surpresa?*](https://www.bcb.gov.br/noticiablogbc/29/noticia){target="_blank"}
  BC Blog, Banco Central do Brasil, post nº 29. — the original post this
  series replicates.
- Alves, S. A. L., & Fasolo, A. M. (2015). *Not Just Another Mixed
  Frequency Paper.* BCB Working Paper 400.
  [PDF](https://www.bcb.gov.br/pec/wps/ingl/wps400.pdf){target="_blank"} —
  source of the retropolated-unemployment-rate methodology this post's
  PNAD series follows.
- Araujo, F., Rodrigues Neto, J. A., & Areosa, M. B. M. (2003). *r-filters:
  a Hodrick-Prescott Filter Generalization.* BCB Working Paper 69. — the
  "r-filter of order 1" state-space representation used above.
- Ball, L., & Mankiw, N. G. (2002). The NAIRU in Theory and Practice.
  *Journal of Economic Perspectives*, 16(4), 115–136. — the sequential
  (non-Kalman) NAIRU approach that motivated the simultaneous MLE
  alternative used here.
